Welcome Home --- Now Vote!\thanks{Prepared for the 2020 Annual Meeting of the Midwest Political Science Association. The author thanks Jacob Faber, Jeff Manza, Myrna Pérez, Ariel White, and Peter Miller for their comments on this project. All errors are my responsibility.}

### 2020-03-04 markdown/part2/part2_standalone.tex
write

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Welcome Home --- Now Vote!},
  pdfauthor={Kevin Morris},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{rotating}
\newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{A\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{A\arabic{figure}}}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {\setlength{\parindent}{0pt}%
  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
  {\par}

\title{Welcome Home --- Now Vote!\thanks{Prepared for the 2020 Annual Meeting of the Midwest Political Science Association. The author thanks Jacob Faber, Jeff Manza, Myrna Pérez, Ariel White, and Peter Miller for their comments on this project. All errors are my responsibility.}}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{Voting Rights Restoration and Post-Supervision Participation}
\author{Kevin Morris\footnote{Researcher, Brennan Center for Justice at NYU School of Law, 120 Broadway Ste 1750, New York, NY 10271 (\href{mailto:kevin.morris@nyu.edu}{\nolinkurl{kevin.morris@nyu.edu}})}}
\date{January 23, 2020}

\begin{document}
\maketitle
\begin{abstract}
This paper presents causal estimates of the effect of voting rights restoration prior to discharge from parole on post-supervision participation. In 2018, New York State began restoring voting rights to parolees, after previously restoring voting rights only at the completion of parole. By leveraging randomness in parole discharge date, I interrogate whether restoring voting rights to parolees increases their post-supervision propensity to cast a ballot. I demonstrate that individuals whose rights were restored while still on parole were more likely to participate than those whose rights were automatically restored upon completion of parole. This group-level effect, however, masks race-specific effects. Although rights restoration prior to parole discharge effectively doubled turnout among white former parolees, it had no measurable effect on the turnout of non-white former parolees. This raises serious questions about how rights restoration programs are implemented, and how incarceration might differently structure black Americans' view of the democratic process.
\end{abstract}

\pagenumbering{gobble}
\pagebreak

@@

In all but two states (Maine and Vermont), felony disenfranchisement laws mean that American citizens convicted of felony offenses lose the right to vote for at least some period of time. In some states, such as Oregon and Massachusetts, individuals lose that right only for the period in which they are actively incarcerated. In other states, notably Kentucky and Iowa, felony convictions result in lifelong disenfranchisement unless a returned citizen receives an individual pardon from the state's governor (Brennan Center for Justice \protect\hyperlink{ref-bcj_laws}{2018}). This variation in laws arises from language in the Fourteenth Amendment which allows states to revoke individuals' voting rights ``for participation in rebellion, or other crime.'' The definition of ``other crime,'' left so vague in the Constitution, is now generally used by states to disenfranchise citizens for any felony offense. The Supreme Court, in cases such as \emph{Richardson v. Ramirez} (1974), has upheld states' right to do just that. Collectively, these laws disenfranchise as many as 4.7 million American citizens. Of these, the majority are no longer incarcerated, but are living and working in their communities (Uggen, Larson, and Shannon \protect\hyperlink{ref-sentencing_2016}{2016}).\footnote{The figures reported in Uggen, Larson, and Shannon (\protect\hyperlink{ref-sentencing_2016}{2016}) have been adjusted to reflect the impact of Amendment 4 in Florida.}

@@

On April 18\textsuperscript{th}, 2018, Governor Andrew Cuomo signed Executive Order 181 which effectively ended the disenfranchisement of New Yorkers on parole. Such a move was of course good for individuals who were still on parole on election day, and would have been disenfranchised otherwise. The change in policy is also beneficial for felony probationers: despite the fact that probationers do not formally lose their voting rights, there is evidence that confusion around the law contributes to \emph{de facto} disenfranchisement among probationers (Drucker and Barreras \protect\hyperlink{ref-Drucker2005}{2005}). The executive order is a promising step: by changing the policy to allow all New York citizens living in their communities to cast a ballot, the move has the potential to both re-enfranchise the nearly 30,000 New Yorkers on parole living in the community and to clarify the rules about who is eligible to vote.

@@

In the analysis below, I examine the effect of Executive Order 181 on individuals who finished parole before October 10\textsuperscript{th}, 2018 (the registration deadline for 2018). These individuals, because they were discharged from parole before the registration deadline, would have been eligible to vote even in the absense of the executive order. This is only a subset of the individuals who were impacted by Executive Order 181 --- many of the individuals who had their voting rights restored were still on parole on election day, and therefore were eligible to vote only because of the rules change. However, by focusing on individuals who would have been eligible to vote either way, I can test whether rights restoration prior to parole discharge serves as an effective encouragement of post-supervision participation.

@@

In the aftermath of the 2000 presidential election, academic interest in the political implications of felony disenfranchisement was stirred thanks to a paper from Uggen and Manza (\protect\hyperlink{ref-Uggen2002}{2002}). George W. Bush's margin of victory in Florida in 2000 was famously just 537 votes. In their 2002 paper, Uggen and Manza estimate the likely partisan composition of the disenfranchised population with felony convictions in their past. They estimate that if this group had been allowed to vote they would have supported Al Gore by a wide margin. Their enfranchisement, Uggen and Manza argued, would have tipped the presidential contest and resulted in the election of Al Gore. They based their estimates on the voting patterns of eligible individuals who were demographically similar to the disenfranchised population. Though much of the research conducted since their 2002 study has pushed back against some of their key assumptions (namely, that formerly incarcerated individuals turn out at the same rate as demographically-similar individuals who have not been incarcerated), Uggen and Manza convincingly demonstrated that felony disenfranchisement can have material political consequences. In the years after Uggen and Manza published their paper, scholars sought to investigate the number of individuals who would have voted if they were not disenfranchised, often using survey data or interviews to construct their estimates (Uggen and Manza \protect\hyperlink{ref-Uggen2004}{2004}; Drucker and Barreras \protect\hyperlink{ref-Drucker2005}{2005}).

In a series of papers between 2009 and 2011, researchers developed methods for directly estimating the turnout of formerly disenfranchised individuals. Haselswerdt (\protect\hyperlink{ref-Haselswerdt2009}{2009}) matched administrative release data and voter registration data from Erie County, NY, to estimate turnout among a small group of formerly incarcerated individuals. Traci Burch expanded upon this matching methodology to estimate the voting patterns of formerly disenfranchised individuals in a range of states (\protect\hyperlink{ref-Burch2010}{2010}, \protect\hyperlink{ref-Burch2011}{2011}). She used release data from states' Departments of Corrections and their registered voter files to identify formerly incarcerated individuals who went on to register to vote. Using the registered voter files, she estimated the party affiliation of formerly incarcerated individuals (in states with party registration) and their turnout rates. Her methodology has been used to investigate other questions surrounding the voting patterns of formerly incarcerated individuals under different circumstances and to examine the impact of changes in disenfranchisement policy (e.g.~Meredith and Morse \protect\hyperlink{ref-Meredith2013}{2013}, \protect\hyperlink{ref-Meredith2015}{2015}).

@@

@@

The literature on turnout among formerly incarcerated individuals leads me to hypothesize that rights restoration prior to parole discharge will boost turnout among formerly incarcerated individuals. I expect that this increased propensity to vote will arise due both thanks to better clarity about individual eligibility to participate, but also because rights restoration may repair some of the distrust of the state engendered by incarceration.

Formerly convicted individuals are very often confused about their eligibility to vote (Drucker and Barreras \protect\hyperlink{ref-Drucker2005}{2005}; Manza and Uggen \protect\hyperlink{ref-locked_out}{2006}). Some research indicates that dispelling misinformation boosts turnout among the formerly disenfranchised. Meredith and Morse (\protect\hyperlink{ref-Meredith2015}{2015}) examines the impact of ending permanent disenfranchisement in Iowa. They find that individuals who received letters explicitly informing them of their re-enfranchisement were more likely to cast ballots in the next election than those who did not. Meredith and Morse (\protect\hyperlink{ref-Meredith2013}{2013}), however, examines states where so-called notification laws went into effect. Although rules about eligibility did not change in these states, new policies required Departments of Corrections to notify formerly disenfranchised individuals of their re-instated voting rights. Meredith and Morse (\protect\hyperlink{ref-Meredith2013}{2013}) finds no effect on turnout from notification in the absence of eligibility changes. Gerber et al. (\protect\hyperlink{ref-Gerber2014}{2014}) conducted a field experiment in Connecticut in advance of the 2012 presidential election, finding that sending mailers to individuals to remind them of their voting rights was successful at increasing turnout among this population. ``Whatever the participatory consequences of incarceration,'' they conclude, ``they are not in large part impossible to overcome'' (924). Even if incarceration does not decrease individuals' propensity to vote, there is reason to believe that reminding formerly incarcerated individuals of their rights increases participation.

Research also indicates that individuals who have negative interactions with the state are less likely to participate in civic life (Pierson \protect\hyperlink{ref-Pierson1993}{1993}). Weaver and Lerman (\protect\hyperlink{ref-Weaver2010}{2010}) argues that ``contact with the institutions of criminal justice is important in structuring patterns of participation long assumed in the dominant literature to stem primarily from aspects of the individual'' (829). To the extent that parole officers are accurately informing their parolees of their newly restored voting rights, Executive Order 181 is likely to bring about a positive interaction between the parolee and the government. Rather than simply have one's rights restored upon completion of sentence, Executive Order 181 may lead parolees to feel individually invited back into the democratic process --- an invitation that may be successful at repairing some of the negative associations created through incarceration. Such repairs may lead to increased propensity to participate in elections in the future.

@@

Following Executive Order 181, the Department of Corrections and Community Supervision began indicating on their online Parolee Lookup Tool whether a parolee had her voting rights restored. By using the identification number provided from the parolee public records request and this website, I was able to identify individuals who had their voting rights restored.\footnote{Not all parolees listed in the public records request data are included in the lookup tool. For individuals who finished parole between January 1\textsuperscript{st}, 2018, and April 17\textsuperscript{th}, 2018, 1.0 percent are not in the lookup tool. For those discharged from parole between April 18\textsuperscript{th}, 2018, and January 13\textsuperscript{th}, 2019 (the latest date of the parole records), 1.2 percent of individuals are not found in the lookup tool.} There were 5,024 individuals who were discharged from parole before the registration deadline whose rights were restored while still under supervision.

@@

\begin{table}[H]

\caption{\label{tab:shift-dobs-chunk}\label{tab:change-dobs} Results of Shifting Birthdates}
\centering
\begin{tabular}[t]{cc}
\toprule
Group & \makecell[l]{Number of Matches Between\\DOCCS and Voter File Records}\\
\midrule
Actual Birthdate & 69,644\\
Birthdate + 35 Days & 502\\
Birthdate - 35 Days & 426\\
\bottomrule
\end{tabular}
\end{table}

@@

@@

Figure \ref{fig:to-18} plots month of parole discharge and turnout in the 2018 midterm elections. Once again, a weighted quadratic curve is fitted with a 95 percent confidence band. This curve is fit on monthly data running from January 2012 through April 2018, and extended through October 2018.

@@

However, the relationship between time-off-parole and propensity to vote is far weaker in the short term. Figure \ref{fig:to-18-recent} indicates that parole discharge date and turnout rates in the 2018 midterm election are not correlated for individuals discharged in 2017 or 2018 before the executive order was enacted.

{\centering \includegraphics{part2_standalone_files/figure-latex/to-18-recent-chart-1}

\caption{\label{fig:to-18-recent}Turnout in 2018 Midterm Election}\label{fig:to-18-recent-chart}
\end{figure}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Welcome Home --- Now Vote!},
  pdfauthor={Kevin Morris},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{rotating}
\newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{A\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{A\arabic{figure}}}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {\setlength{\parindent}{0pt}%
  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
  {\par}

\title{Welcome Home --- Now Vote!\thanks{Prepared for the 2020 Annual Meeting of the Midwest Political Science Association. The author thanks Jacob Faber, Jeff Manza, Myrna Pérez, Ariel White, and Peter Miller for their comments on this project. All errors are my responsibility.}}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{Voting Rights Restoration and Post-Supervision Participation}
\author{Kevin Morris\footnote{Researcher, Brennan Center for Justice at NYU School of Law, 120 Broadway Ste 1750, New York, NY 10271 (\href{mailto:kevin.morris@nyu.edu}{\nolinkurl{kevin.morris@nyu.edu}})}}
\date{March 04, 2020}

\begin{document}
\maketitle
\begin{abstract}
This paper presents causal estimates of the effect of voting rights restoration prior to discharge from parole on post-supervision participation. In 2018, New York State began restoring voting rights to parolees, after previously restoring voting rights only at the completion of parole. By leveraging randomness in parole discharge date, I interrogate whether restoring voting rights to parolees increases their post-supervision propensity to cast a ballot. I demonstrate that individuals whose rights were restored while still on parole were more likely to participate than those whose rights were automatically restored upon completion of parole. This group-level effect, however, masks race-specific effects. Although rights restoration prior to parole discharge effectively doubled turnout among white former parolees, it had no measurable effect on the turnout of non-white former parolees. This raises serious questions about how rights restoration programs are implemented, and how incarceration might differently structure Black Americans' view of the democratic process.
\end{abstract}

\pagenumbering{gobble}
\pagebreak
\doublespacing

@@

In all but two states (Maine and Vermont), felony disenfranchisement laws mean that American citizens convicted of felony offenses lose the right to vote for at least some period of time. In some states, such as Oregon and Massachusetts, individuals lose that right only for the period in which they are actively incarcerated. In other states, notably Kentucky and Iowa, felony convictions result in lifelong disenfranchisement unless a returned citizen receives an individual pardon from the state's governor (Justice \protect\hyperlink{ref-bcj_laws}{2019}). This variation in laws arises from language in the Fourteenth Amendment which allows states to revoke individuals' voting rights ``for participation in rebellion, or other crime.'' The definition of ``other crime,'' left so vague in the Constitution, is now generally used by states to disenfranchise citizens for any felony offense. The Supreme Court, in cases such as \emph{Richardson v. Ramirez} (1974), has upheld this practice. Collectively, these laws disenfranchise as many as 4.7 million American citizens. Of these, the majority are no longer incarcerated(Uggen, Larson, and Shannon \protect\hyperlink{ref-sentencing_2016}{2016}).\footnote{The figures reported in Uggen, Larson, and Shannon (\protect\hyperlink{ref-sentencing_2016}{2016}) have been adjusted to reflect the impact of Amendment 4 in Florida.}

@@

In the spring of 2018,\footnote{Although the executive order was signed on April 18\textsuperscript{th}, it did not actually go into effect until May 18\textsuperscript{th}.} Governor Andrew Cuomo signed Executive Order 181 which effectively ended the disenfranchisement of New Yorkers on parole. Such a move was of course good for individuals who were still on parole on election day, and would have been disenfranchised otherwise. The change in policy is also beneficial for felony probationers: despite the fact that probationers do not formally lose their voting rights, there is evidence that confusion around the law contributes to \emph{de facto} disenfranchisement among probationers (Drucker and Barreras \protect\hyperlink{ref-Drucker2005}{2005}). The executive order is a promising step: by changing the policy to allow all New York citizens living in their communities to cast a ballot, the move has the potential to both re-enfranchise the nearly 30,000 New Yorkers on parole living in the community and to clarify the rules about who is eligible to vote.

@@

In the analysis below, I examine the effect of Executive Order 181 on individuals who finished parole before October 10\textsuperscript{th}, 2018 (the registration deadline for 2018). These individuals, because they were discharged from parole before the registration deadline, would have been eligible to vote even in the absence of the executive order. This is only a subset of the individuals who were impacted by Executive Order 181 --- many of the individuals who had their voting rights restored were still on parole on election day, and therefore were eligible to vote only because of the rules change. However, by focusing on individuals who would have been eligible to vote either way, I can test whether rights restoration prior to parole discharge serves as an effective encouragement of post-supervision participation.

By examining the effect of rights restoration in the context of a personal relationship with a parole officer, rather than simply at the end of a sentence and communicated through the mail, Executive Order 181 offers us the opportunity to build on existing research. Do these human interactions repair damage done to the formerly incarcerated individual's relationship to the state? Or do parole officers exhibit biases against their stewards, making them ineffective conduits for restoration?

@@

In a series of papers between 2009 and 2011, researchers developed methods for directly estimating the turnout of formerly disenfranchised individuals. Haselswerdt (\protect\hyperlink{ref-Haselswerdt2009}{2009}) matched administrative release data and voter registration data from Erie County, NY, to estimate turnout among a small group of formerly incarcerated individuals. Traci Burch expanded upon this matching methodology to estimate the voting patterns of formerly disenfranchised individuals in a range of states (\protect\hyperlink{ref-Burch2012}{2012}, \protect\hyperlink{ref-Burch2011}{2011}). She used release data from states' Departments of Corrections and their registered voter files to identify formerly incarcerated individuals who went on to register to vote. Using the registered voter files, she estimated the party affiliation of formerly incarcerated individuals (in states with party registration) and their turnout rates. Her methodology has been used to investigate other questions surrounding the voting patterns of formerly incarcerated individuals under different circumstances and to examine the impact of changes in disenfranchisement policy (e.g.~Meredith and Morse \protect\hyperlink{ref-Meredith2013}{2013}, \protect\hyperlink{ref-Meredith2015}{2015}).

@@

\hypertarget{notification-re-enfranchisement-and-turnout}{%
\section*{Notification, Re-enfranchisement, and Turnout}\label{notification-re-enfranchisement-and-turnout}}
\addcontentsline{toc}{section}{Notification, Re-enfranchisement, and Turnout}

@@

Formerly convicted individuals are very often confused about their eligibility to vote (Drucker and Barreras \protect\hyperlink{ref-Drucker2005}{2005}; Manza and Uggen \protect\hyperlink{ref-locked_out}{2008}). Some research indicates that dispelling misinformation boosts turnout among the formerly disenfranchised. Meredith and Morse (\protect\hyperlink{ref-Meredith2015}{2015}) examines the impact of ending permanent disenfranchisement in Iowa. Prior to 2005 (and after 2011), individuals with felony convictions were permanently disenfranchised unless they submitted an application to the governor. Executive Order 42 eliminated this requirement, instead re-enfranchising automatically upon being discharged from their sentence. Although all formerly disenfranchised individuals were re-enfranchised upon the signing of the Executive Order, not all re-enfranchised individuals were informed of their change in status. Meredith and Morse (\protect\hyperlink{ref-Meredith2015}{2015}) leverages differences in notification to test whether the notification increased turnout, finding a strong positive effect. Meredith and Morse (\protect\hyperlink{ref-Meredith2013}{2013}), however, examines states where so-called notification laws went into effect. Although rules about eligibility did not change in these states, new policies required Departments of Corrections to notify formerly disenfranchised individuals of their re-instated voting rights. Meredith and Morse (\protect\hyperlink{ref-Meredith2013}{2013}) finds no effect on turnout from notification in the absence of eligibility changes.

Gerber et al. (\protect\hyperlink{ref-Gerber2015}{2015}) built on the quasi-experimental design of Meredith and Morse (\protect\hyperlink{ref-Meredith2015}{2015}) and conducted a field experiment in Connecticut in advance of the 2012 presidential election. In their field experiment, some formerly convicted (but eligible) residents were reminded of their eligibility; others were not. Like Meredith and Morse (\protect\hyperlink{ref-Meredith2015}{2015}), they find evidence that these reminders successfully increased turnout. ``Whatever the participatory consequences of incarceration,'' they conclude, ``they are not in large part impossible to overcome'' (924). Even if incarceration does not decrease individuals' propensity to vote, there is reason to believe that reminding formerly incarcerated individuals of their rights increases participation.

Meredith and Morse (\protect\hyperlink{ref-Meredith2015}{2015}) and Gerber et al. (\protect\hyperlink{ref-Gerber2015}{2015}) provide valuable insight into mechanisms that can increase the turnout of formerly disenfranchised individuals. Nevertheless, the research to date has examined the effect of mail notification on post-supervision turnout. In the case of New York State, formerly incarcerated individuals were being informed of their voting rights through the mail when they finished their sentence prior to the executive order. This study tests, therefore, not the efficacy of notification laws but rather whether in-person restoration prior to discharge increases turnout above-and-beyond any increase associated with mail notification.

There is reason to believe that some of the negative treatment effects of incarceration arise from the very nature of person-to-person relationships in the criminal justice system. Lerman and Weaver (\protect\hyperlink{ref-Lerman2014}{2014}) argues that contact with the criminal justice system can uniquely structure democratic participation more than other types of government contact. ``It may also be,'' they write, ``that social benefits {[}arising from non-criminal justice contact with the state{]} are less visible to citizens because they often occur through the mail, rather than through personal contact with government agencies or officials, making it easier to disconnect social benefits from government and the political system'' (93). Because Executive Order 181 operates \emph{not} through the mail but rather through ``personal contact with government\ldots{} officials,'' its effects may differ from the mail notification programs. The social benefits (restoration of voting rights) might be communicated more effectively, thus increasing turnout further.

Of course, reliance on a human-mediated system to deliver social benefits poses potential practical and scientific complications. A notification sent in the mail is a binary treatment: the letter was either sent or not sent. Communicating the restoration of voting rights through human interaction means that the treatment might vary by individuals: if parolees have a strong relationship with their parole officer, they may be strongly encouraged to register and participate. On the other hand, if parole officers are biased against their parolees (due either to racial hostility, personal dislike, or expected partisan affiliation) they may not provide any effective treatment at all. Thus, while the communication of voting rights restoration in the context of a personal relationship has the potential to deliver much larger benefits, it also opens the door to potential discrepancies in treatment.

I expect that restoring voting rights while an individual is still on parole increases that individual's later propensity to vote substantially. This increase will operate through two mechanisms: as the previous literature has established, notification of voting rights can increase turnout, and being informed by a parole officer likely leaves a former parolee more sure of her voting rights than a letter in the mail. Secondly, the interpersonal nature of the parolee-parole officer relationship is expected to more effectively repair (some of) the damage done to the former parolees interpretation of the state. The individual, in-person invitation to rejoin the body politic was likely effective.

@@

Following Executive Order 181, the Department of Corrections and Community Supervision began indicating on their online Parolee Lookup Tool whether a parolee had her voting rights restored. By using the identification number provided from the parolee public records request and this website, I was able to identify individuals who had their voting rights restored.\footnote{Not all parolees listed in the public records request data are included in the lookup tool. Roughly 1 percent of former parolees are not in the online lookup.} There were 3,093 individuals who were discharged from parole before the registration deadline whose rights were restored while still under supervision.

@@

\begin{singlespace}
\begin{table}[H]

\caption{\label{tab:shift-dobs-chunk}\label{tab:change-dobs} Results of Shifting Birthdates}
\centering
\begin{tabular}[t]{cc}
\toprule
Group & \makecell[l]{Number of Matches Between\\DOCCS and Voter File Records}\\
\midrule
Actual Birthdate & 69,644\\
Birthdate + 35 Days & 502\\
Birthdate - 35 Days & 426\\
\bottomrule
\end{tabular}
\end{table}
\end{singlespace}

@@

\hypertarget{research-design}{%
\section*{Research Design}\label{research-design}}
\addcontentsline{toc}{section}{Research Design}

In this paper, I exploit randomness in discharge date from parole in New York State to determine the efficacy of the policy change. All individuals discharged from parole after the start of the executive order are assigned to the ``treatment'' group; individuals discharged earlier are assigned to the control group. When I limit the pool of formerly incarcerated individuals to those discharged from parole in 2017 and 2018, the control and treatment voters look virtually identical. The assignment to treatment, therefore, is as nearly random as can be hoped for in the context of a natural experiment.

The analysis is complicated, however, by the fact that not all individuals discharged from parole after the policy change --- that is to say, assigned to the treatment group --- were actually ``treated'' by having their voting rights restored while on parole in a meeting with their parole officer. In other words, our natural experiment has ``imperfect compliance.'' Noncitizens, for instance, were not eligible for rights ``restoration'' while on parole because in New York State they were not allowed to vote even prior to their conviction. Because not all individuals assigned to the treatment group were in fact treated, simply comparing turnout between the groups discharged on either side of the cutpoint reveals not the treatment effect, but rather the intend-to-treat effect. Put differently, it allows us to estimate how much the executive order increased turnout among former parolees as a group --- including those ineligible to vote --- but not the individual-level effect of the policy change on propensity to vote.

Although we can identify the individuals in the assigned-to-treatment group who do not ``comply'' with their treatment status, we do not know who among the ``assigned-to-control'' group would have gone untreated if they had been assigned to the other group. We cannot identify, for instance, noncitizens who were discharged from parole before the policy change. In an ideal world, we would limit both the treatment \emph{and} the control group to individuals who were eligible to have their rights restored while on parole; we do not live in such a world. Worse yet, there is obviously a relationship between compliance (rights restoration) and the dependent variable of interest: noncitizens are, of course, both ineligible to have voting rights bestowed and cannot vote.

The standard way of dealing with imperfect compliance in an experimental setting is to leverage the (random) assignment to treatment as an instrument for the (nonrandom) actual compliance (Angrist, Imbens, and Rubin \protect\hyperlink{ref-Angrist1996}{1996}). Such an approach has been widely used in the context of voter turnout (e.g.~Ansolabehere, Iyengar, and Simon \protect\hyperlink{ref-Ansolabehere1999}{1999}; Gerber and Green \protect\hyperlink{ref-Gerber2000}{2000}; Milligan, Moretti, and Oreopoulos \protect\hyperlink{ref-Milligan2004}{2004}; Lassen \protect\hyperlink{ref-Lassen2004}{2004}; Sondheimer and Green \protect\hyperlink{ref-Sondheimer2010}{2010}). Gerber and Green (\protect\hyperlink{ref-Gerber2000}{2000}) provides a helpful example of why this two-step process is necessary. The authors test the efficacy of phone calls on increasing turnout. One group is assigned to be the control; the authors do not attempt to call them. Another set of voters receive a phone call, but only some of them are actually contacted, or ``treated'' with the call. The authors could simply test to see whether those contacted by phone turned out at higher rates --- but what if high propensity voters are more likely to answer the phone? Then ``treatment'' would simply be identifying a group of individuals with higher vote propensity than the control group without actually speaking to the efficacy of the treatment.

The same process is likely at play here. The control group includes individuals that \emph{would have} had their rights restored at an in-person meeting if they had been discharged later; it also includes individuals who would not have received the treatment thanks to their citizenship status. By comparing turnout among the actually treated to the control group, we may simply be comparing the turnout of a group of \emph{citizens} to the turnout of a group that includes both citizens \emph{and} noncitizens. Although a positive ``treatment effect'' is likely in this context, it would be unable to speak to causality.

Figure \ref{fig:flow} demonstrates the buckets into which former parolees can fall. Although I argue that a parolee's assignment to the control or treatment categories is random and uncorrelated with turnout except via the treatment, whether a treated former parolee is treated or is a noncomplier is not random and is associated with their propensity to vote through mechanisms other than the treatment.

{\centering \includegraphics{../../temp/flowchart}

\caption{\label{fig:flow}Categorization of Former Parolees}\label{fig:flow}
\end{figure}

Using the natural experiment with imperfect compliance setup, I begin by estimating the intend-to-treat effect of the policy change. This approach simply asks whether, as a whole, turnout was higher among individuals discharged after the policy change than those discharged earlier. I turnout in the 2018 election as the dependent variable in these logistic regressions.

After estimating the intend-to-treat effect of the policy change, I implement the instrumental variables approach to estimate the individual-level effect of in-person voting rights restoration on subsequent turnout. As discussed above, the human nature of the treatment allows for discrepancies in how the treatment is administered. To explore this potential, I also explore whether the treatment had different effects for different subgroups of the formerly incarcerated population. I finally explore whether the treatment effects were larger for individuals who spent more time on parole after having their rights restored.

@@

Figure \ref{fig:to-18} plots month of parole discharge and turnout in the 2018 midterm elections. Once again, a weighted quadratic curve is fitted with a 95 percent confidence band. This curve is fit on monthly data running from January 2012 through May 2018, and extended through October 2018.

@@

However, Table \ref{tab:to-18-logit-short} indicates that this relationship breaks down in the short-run. These logistic models include individuals last discharged from parole between January 1\textsuperscript{st}, 2017, and May 18\textsuperscript{th}, 2018.

\begin{singlespace}

\input{"../../temp/table3.tex"}
\end{singlespace}
